---
title: "Transform camera trap data to recurrent events data"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ctrecurrent}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(knitr)
opts_chunk$set(
  fig.width = 5,
  fig.height = 5,
  fig.align = "center",
  crop      = TRUE)
```

The goal of **`ctrecurrent`** is to transform the camera trap data into a format
suitable for recurrent event analysis.
It contains the function `ct_torecurrent` to do so, requiring a data frame with the
following information for each observation^[If your data is a different format, you can try to bring it into the required format before using the functions provided here, or post an issue on GitHub if you use a common camera trap format, so we can extend the provided functionality.]:

- Site ID,
- Timestamp (Date and Time) and
- Species

## Overview of the raw data
In the first step we look at the `?murphy` data available with the package:

```{r eval = TRUE}
library(dplyr)
library(pammtools)
library(ggplot2)
theme_set(theme_bw())
library(patchwork)
library(reReg)

head(murphy)
unique(murphy$Species)
range(murphy$DateTime)
table(murphy$matrix)
```
As we can see above, the camera captured different types of objects. In
the analysis below, we will focus on "Deer" as the primary species and "Coyote"
as secondary. Everything else will be considered tertiary species.
Thse `matrix` covariate contains information about the environment in which the camera is placed, either agriculturally developed or forest.


## Defining the parameters of the survey
In the next step, we have to define which species we want to consider as the **primary** and **secondary**. This is important, as this dictates the direction of the effect.
Below we define "Deer" as the primary species and "Coyote" as secondary, thus the model will estimate how the presence of Deer affects the time until the occurence of a Coyote.
Occurence of a **tertiary** species will be considered a censoring event.
Finally, we have to set the end date of the study.
The last two arguments do not necessarily need to be specified, the default is to take all species outside of primary and secondary as tertiary and the maximum date in the data as survey end date.

```{r, eval = TRUE}
primary = c("Deer")
secondary = c("Coyote")

# Define the tertiary species (affecting the secondary species, censoring the survey if observed but not being the focus of the model)
tertiary = c("Fawn", "Bear", "Bobcat", "Human", "Motorized")

# Define the end of study
end_date = max(murphy$DateTime)
```

## Transform the raw data for analysis

The data transformation for analysis consists of 2 steps

  1) transform data to general recurrent events formats (function `ct_to_recurrent`).
  Here, at each site, we look for the first occurence of the primary species, which indicates the start of the first survey at this site. Covariate information is added as needed. Occurence of the secondary species will be considered a (recurrent) event within each survey, and the survey at that site continues until
      - occurence of a tertiary species (end of survey -> censoring)
      - end of survey period (`survey_duration` -> administrative censoring)
      - re-occurence of the primary species (start of a new survey)

  2) transform the "piece-wise exponential data" (PED) format for analysis using PAMMs (function `as_ped` from **`pammtools`** package). This transforms the time-to-event data from the recurrence data from step 1. by splitting the time-axis in intervals which facilitates estimation of the hazard for an occurence of the secondary species (see [here](https://adibender.github.io/pammtools/articles/data-transformation.html) and [here](https://adibender.github.io/pammtools/articles/recurrent-events.html))


```{r, eval = TRUE}
# Step 1) transform to recurrent event format
# note that datetime_var, species_var and site_var have defaults, but they need to
# be adjusted if your column names are different
recu = ct_to_recurrent(
  data            = murphy,
  primary         = primary,
  secondary       = secondary,
  tertiary        = tertiary,
  datetime_var    = "DateTime",
  species_var     = "Species",
  site_var        = "Site",
  survey_end_date = end_date,
  survey_duration = 30)
recu |> select(Site, survey_id, t.start, t.stop, event, status, enum)

# Merge covariate information with recurrent events data table
cov = murphy %>%
  select(Site, matrix) %>%
  distinct()
data = left_join(recu, cov)

data |> select(Site, survey_id, primary, secondary, t.start, t.stop, event, matrix)

# Events plot
reReg::plotEvents(
  Recur(t.start %to% t.stop, survey_id, event, status) ~ 1,
  data = data,
  xlab = "Time in days",
  ylab = "Subjects arranged by terminal time")

# By levels of matrix landscape
reReg::plotEvents(
  Recur(t.start %to% t.stop, survey_id, event, status) ~ matrix,
  data = data,
  xlab = "Time in days",
  ylab = "Subjects arranged by terminal time")

# PED transformation
ped = data %>%
  as_ped(
    formula    = Surv(t.start, t.stop, event)~ matrix + Site,
    id         = "survey_id",
    transition = "enum",
    timescale  = "calendar") |>
  mutate(Site = as.factor(Site)) |>
  mutate(survey_id = as.factor(survey_id))

# check the data for one survey
ped |>
  filter(survey_id == "2016BE15-3") |>
  group_by(enum) |>
  slice(1, n()) |>
  select(survey_id, tstart, tend, enum, ped_status)

```

## Fit the model
In the final step, we fit a model to the PED data, which differs depending on the
assumptions about the process. Here we fit 4 models

- `m_null`: baseline model for hazard of Coyote occurence (without covariates)
- `m_ph`: proportional hazards model for the effect of landscape
- `m_tv`: stratified hazards model (each landscape category has its own baseline hazard)
- `m_re`: As `m_tv` but with random effect for camera trap site

```{r, eval = TRUE}
# Baseline model
m_null = pamm(
  formula = ped_status ~ s(tend),
  data = ped)
summary(m_null)

# Proporitional hazards effect of landscape (matrix)
m_ph = pamm(
  formula = ped_status ~ matrix + s(tend),
  data = ped)
summary(m_ph)

 # Time-varying covariate effect
m_tv = pamm(
  formula = ped_status ~ matrix + s(tend) +
     s(tend, by = as.ordered(matrix)),
  data = ped,
  engine = "bam", method = "fREML",  discrete = TRUE)
summary(m_tv)

 m_re <- pamm(
  formula = ped_status ~ matrix + s(tend) + s(tend, by = as.ordered(matrix)) +
    s(Site, bs = "re"),
  data = ped, engine = "bam", method = "fREML", discrete = TRUE
 )
summary(m_re)
```


## Effect visualization
While the models have their own default plot methods, the most flexible way
to visualize non-linear effects is to

- create a new data set with the covariate specification of interest (`make_newdata`)
- add the prediction of the quantity of interest (hazard, hazard ratio, etc.) to your data (`add_*` functions), including CIs
- use your favorite tool for visualization (here we use **`ggplot2`**)

The **`pammtools`** package provides helper functions for the first two steps.

In the first plot, we can see the attraction effect of deer on coyotes
lasts for about 5 days. Afterwards the hazard for the occurence of coyotes is constant (the wigglines can be ignored since the uncertainty is quiet high).
```{r}

 # Null model
ndf_null <- ped %>%
    make_newdata(tend = unique(tend)) %>%
    add_hazard(m_null)
p_null = ggplot(ndf_null, aes(x = tend, y = hazard)) +
   geom_line() +
   geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = .3) +
   xlab("Time after white-tailed deer sighting (in days)") +
   ylab("Log-hazard for coyote occurence")
print(p_null)
```

In the second plot, we plot the hazards for coyote occurence for each category
of landscape (`matrix`) in one plot (left panel) and their hazard ratio(right panel).
This indicates (hazard ratio = 1), that given a sighting of a deer (primary),
occurence of coyotes is not affected by the type of landscape in which the deer was sighted.

```{r, fig.width = 6, fig.height= 3}
# Stratified hazards
ndf_tv <- ped %>%
    make_newdata(tend = unique(tend), matrix = unique(matrix)) %>%
    add_hazard(m_tv)
p_tv = ggplot(ndf_tv, aes(x = tend, y = hazard, colour = matrix)) +
   geom_line() +
   geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = .3) +
   scale_color_manual(values = c("#999999", "#E69F00"))  +
   xlab("Time after white-tailed deer sighting (in days)") +
   ylab("Log-hazard for coyote occurence")
ndf_hr <- ped %>%
    make_newdata(tend = unique(tend), matrix = c("agdev")) %>%
    add_hazard(m_tv, reference = list(matrix = "forest"))
p_hr = ggplot(ndf_hr, aes(x = tend, y = hazard)) +
   geom_line() +
   geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = .3) +
   xlab("Time after white-tailed deer sighting (in days)") +
   ylab("Hazard ratio for coyote occurence\n (agdev/forest")
p_tv + p_hr
```
